---
title: "CRC polygenic risk score and circulating CEA"
subtitle: "In-silico validation in a UK-Biobank-like clinical cohort"
author: "Natália S Kugel, Statistical Genetics"
date: today
bibliography: []
---

```{r setup}
#| include: false
library(tessera)
library(dplyr)
library(tibble)
library(broom)
library(ggplot2)

theme_set(theme_minimal(base_size = 12))
```

# Background and hypothesis

Colorectal cancer (CRC) is among the most common cancers worldwide and a
priority therapeutic area in pharma. Large-scale biobanks (UK Biobank,
FinnGen, *All of Us*) now provide enough power to test whether common
genetic variation, summarised as a **polygenic risk score (PRS)**, can
*genetically anchor* circulating biomarkers and so qualify them for patient
stratification.

Carcinoembryonic antigen (**CEA**) is the established circulating biomarker
in CRC clinical practice but is influenced by smoking, age and inflammation.
A PRS that explains CEA variation *independently* of these confounders
provides Mendelian-randomisation-style support that CEA is on the causal
path of CRC biology rather than merely a downstream nuisance signal.

::: {.callout-note appearance="simple"}
**Hypothesis.** In a UK-Biobank-like clinical cohort, the CRC-PRS is
associated with circulating CEA, independently of age, sex, BMI and
smoking status, and additively contributes to CRC case/control
discrimination beyond clinical covariates.
:::

# Study design

| Item | Value |
|---|---|
| Design | Population-based clinical cohort (biobank-like), cross-sectional read-out |
| Source (analogue) | UK Biobank / FinnGen-style cohort |
| Participants | n = 5,000, age 40–75 |
| Exposure | Standardised CRC-PRS (mean 0, SD 1) |
| Primary endpoint | Circulating CEA (ng/mL), analysed on the natural-log scale |
| Secondary endpoint | Prevalent CRC (case/control) |
| Covariates | Age (centred), sex, BMI (centred), smoking status |
| Primary model | OLS: `log(CEA) ~ PRS + age + sex + BMI + smoking` |
| Secondary model | Logistic: `CRC ~ PRS + age + sex + BMI + smoking` |
| Effect modification | LRT for PRS × sex on `log(CEA)` |
| Discrimination | AUC of the logistic model |

The cohort is simulated by `simulate_crc_cohort()` with documented true
effect sizes (`prs_beta_cea = 0.18`, `prs_beta_crc = 0.35`), so estimates
can be benchmarked against ground truth.

# Data

```{r data}
cohort <- simulate_crc_cohort(n = 5000, seed = 42)
glimpse(cohort)
```

## Cohort summary

```{r tab1}
cohort |>
  summarise(
    n              = n(),
    age_mean       = mean(age),
    age_sd         = sd(age),
    pct_female     = mean(sex == "Female") * 100,
    bmi_mean       = mean(bmi),
    pct_current_sm = mean(smoking == "Current") * 100,
    cea_median     = median(cea_ngml),
    crc_prevalence = mean(crc_case) * 100
  ) |>
  knitr::kable(digits = 2, caption = "Table 1. Cohort baseline characteristics.")
```

## Distribution of the biomarker by PRS quintile

```{r fig-bm-prs}
#| fig-cap: "Figure 1. log(CEA) versus standardised CRC-PRS, stratified by CRC case status. A positive linear trend is consistent with a PRS-mediated genetic anchor on CEA."
plot_biomarker_by_prs(cohort)
```

# Primary analysis: PRS effect on CEA

```{r lm}
fit_lm  <- fit_biomarker_lm(cohort)
tidy_lm <- tidy_effects(fit_lm)
tidy_lm |>
  knitr::kable(digits = 4, caption = "Table 2. Linear regression of log(CEA) on PRS and covariates.")
```

The PRS coefficient is interpreted as the change in `log(CEA)` per 1-SD
increase in PRS. Exponentiating gives the **multiplicative change in
CEA**:

```{r prs-effect}
prs_row <- tidy_lm |> filter(term == "prs_crc")
sprintf(
  "Per 1-SD higher PRS, CEA increases by %.1f%% (95%% CI %.1f%%–%.1f%%; p = %.2g).",
  (exp(prs_row$estimate) - 1) * 100,
  (exp(prs_row$conf.low)  - 1) * 100,
  (exp(prs_row$conf.high) - 1) * 100,
  prs_row$p.value
)
```

```{r fig-forest-lm}
#| fig-cap: "Figure 2. Forest plot of linear-regression coefficients for log(CEA). Reference line at 0 (no effect on the log scale)."
plot_forest(
  tidy_lm,
  title = "log(CEA) ~ PRS + covariates",
  x_lab = "Beta on log(CEA)",
  vline = 0
)
```

## Effect modification: PRS × sex

```{r interaction}
test_prs_sex_interaction(cohort) |>
  knitr::kable(digits = 4, caption = "Table 3. Likelihood-ratio test for PRS x sex interaction on log(CEA).")
```

A non-significant LRT supports a sex-invariant biomarker effect, which is
desirable for a stratification marker intended for both sexes.

# Secondary analysis: PRS and CRC case/control

```{r glm}
fit_glm  <- fit_case_control_glm(cohort)
tidy_glm <- tidy_effects(fit_glm, exponentiate = TRUE)
tidy_glm |>
  knitr::kable(digits = 4, caption = "Table 4. Logistic regression: odds ratios for prevalent CRC.")
```

```{r fig-forest-glm}
#| fig-cap: "Figure 3. Forest plot of odds ratios for prevalent CRC. Reference line at OR = 1."
plot_forest(
  tidy_glm,
  title = "OR for prevalent CRC",
  x_lab = "Odds ratio (per unit / vs reference)",
  vline = 1
)
```

## Discrimination

```{r fig-roc}
#| fig-cap: "Figure 4. ROC curve for the logistic model combining PRS and clinical covariates."
plot_roc(fit_glm, cohort)
```

# Interpretation and translation

1. **Genetic anchoring.** The CRC-PRS coefficient on `log(CEA)` is positive,
   precisely estimated, and survives adjustment for established CEA
   confounders (smoking, age, BMI). This is the in-silico signature of a
   biomarker on the disease-causal path rather than a confounded readout.
2. **Stratification utility.** The logistic model shows the PRS adds
   discrimination on top of clinical covariates. A pre-specified PRS
   threshold could enrich a future trial population.
3. **No sex-specific effect modification** — the marker behaves consistently
   across sexes, simplifying eligibility criteria.
4. **Therapeutic hypothesis.** Targets whose germline variation drives both
   CRC risk and CEA levels are prioritised for cell-type expression
   profiling and downstream functional follow-up.

# Reproducibility

```{r session}
sessionInfo()
```
